Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-15 writeup/conStruct.tex
moved over to plos template

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{float}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[authoryear]{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bigints}
\usepackage{array}
\usepackage{tikz}
\usepackage{longtable}
\usepackage{geometry}

\bibliographystyle{plainnat}

\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0pt}

@@

\geometry{a4paper}

\title{Inferring Continuous and Discrete Population Genetic Structure Across Space}
\date{\vspace{-5ex}}
\author{
Gideon S.\ Bradburd$^{1,a}$, 
Graham M.\ Coop$^{2,b}$,
Peter L.\ Ralph$^{3,c}$}

\begin{document}

\maketitle

\textsuperscript{1}
Department of Integrative Biology, 
Ecology, Evolutionary Biology, and Behavior Graduate Group,
Michigan State University, MI 48824

\textsuperscript{2}
Center for Population Biology,
Department of Evolution and Ecology, 
University of California, Davis, CA 95616

\textsuperscript{3}
Institute of Ecology and Evolution,
Department of Mathematics,
University of Oregon, Eugene, OR 97403

\textsuperscript{a}bradburd@msu.edu; 
\textsuperscript{b}gmcoop@ucdavis.edu;
\textsuperscript{c}plr@uoregon.edu\\\\\

\newpage

\begin{abstract}
A classic problem in population genetics is the characterization 
of discrete population structure in the presence of 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
the use of clustering or assignment methods may incorrectly ascribe
differentiation due to continuous processes (e.g., geographic isolation by distance)
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
% This is partly a result of the difficulty of sampling uniformly and continuously 
% across the range of a population or species, but more, it 
This reflects a shortcoming of current methods for inferring and 
visualizing population structure when applied to genetic data
deriving from geographically distributed populations.
Here, we present a novel statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of two-dimensional population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This model explicitly addresses the ``clines vs. clusters" problem in 
modeling population genetic variation.
% by jointly accommodating both continuous and discrete patterns of differentiation. 
The method produces useful descriptions of structure in genetic relatedness
in situations where separated, geographically distributed populations interact,
as after a range expansion or secondary contact.
We demonstrate the utility of this approach using simulations 
and by applying it to empirical datasets of poplars and black bears in North America.
\end{abstract}

@@

\section*{Methods}

@@

This definition of covariance differs from the usual ``genetic covariance'' \citep{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eqn. \eqref{allelic_covariance} further in the Appendix (\secref{allelic_cov}).
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $\pi_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 \pi_{i,j})/4$.
Therefore, 
this allelic covariance is more affected in shape by singleton sites
than the standard genetic covariance,
so it may be advisable to filter these prior to analysis
if they are likely to contain a large percentage of errors.

@@

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each layer, 
we define the covariance within layer $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_layer_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th layer.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
and the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the layer.
Our choice of a powered-exponential decay, 
as parameterized by the $\alpha$s, 
is a flexible and standard choice in spatial statistics \citep{Diggle1998}, 
and is not chosen to match a particular population genetics model. 
% (Analytical theory \citep{barton2002neutral} predicts the decay should be proportional to $\exp{-D}/\sqrt{D}$.)
The $\phi^{(k)}$ is a parameter that describes the background covariance within the layer. 
If two samples draw 100\% of their ancestry from layer $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the layer.
The ``shared drift'' parameter $\phi^{(k)}$ is analogous to 
the branch length connecting the $k$th population to the population ancestral to all modeled
layers \citep[see for example ][]{patterson_ancient_2012,peter_fstats},
although they cannot be directly compared because 
we are modeling the allelic, rather than genetic, covariance. 
In the Appendix section \ref{rationale} we lay out a
simple model of allele frequencies underlying this covariance model.

@@

A nice feature of our approach is that the model described in Eqn \eqref{cross_layer_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} in the Appendix for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is very similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g.][]{ADMIXTURE}; 
the \gc{main} difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomially distributed \citep{Engelhardt2012}.  
(We make this approximation for the substantial advantages in
computational speed.)
\gc{The second small difference is that in the original STRUCTURE model,
  allele frequencies at each locus are independently drawn for each
  cluster \citep{STRUCTURE}, while in conStruct's non-spatial model,
  it is more natural to envision each layer's allele frequency
  as being drifted away from a single, global allele frequency. 
  This makes our model more
  closely related to the ``$F$-model'' prior for allele frequencies of \citet{falush2003}.
As these differences are relatively small,} we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework. 
We also validate our claim that the results of our nonspatial method 
are a close match to those of STRUCTURE-based approaches.

@@

We use cross-validation \citep[similar in spirit to][]{ADMIXTURE_xval} to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} in the Appendix for more on our cross-validation procedure).
In general, larger values of $K$ allows the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
since above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

@@

To calculate layer contributions, we use the following alternative description of our covariance model:
the genomes of any pair of individuals \emph{agree} with some background probability,
but this probability of agreement is increased on any segment of genome 
that both have inherited through the same layer 
(the amount it increases depends on how far apart they are.)
We use this characterization to quantify the relative contributions of each layer,
by computing the average contribution to increased probability of agreement
as described in Appendix \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

\section*{Results}

@@

To test the method, we first generated data using the coalescent simulator \texttt{ms} \citep{Hudson2002}.
In each simulation,
we split a single ancestral population into $K$ subpopulations $\tau_{\text{s}}$ units of coalescent time in the past,
and at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonized a
separate $6 \times 6$ square lattice of demes.
Migration on each lattice was to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past, 
we collapsed those $K$ discrete layers into a single grid of demes,
choosing various amounts of admixture from these different layers
(see Fig \ref{sim_setup}). 
We collapsed the layers together using random admixture with spatially autocorrelated proportions 
(see Appendix \secref{sim_details} for details), 
so that both the frequencies in layers and the layer memberships are
spatially autocorrelated. 
We simulated datasets using $K=1$, 2, and 3 layers;
in each simulation we sampled 10,000 unlinked loci from each of 20 haploid individuals from every deme.
We then ran both spatial and nonspatial \texttt{conStruct} analyses on each simulated dataset with $K$ between 1 and 7,
and compared predictive performance of the models using cross-validation.
% To confirm that results from our nonspatial model were qualitatively similar to those of 
% other nonspatial model-based clustering algorithms, 
For comparison,
we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, using $K=2-4$.
% inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}--\ref{fastStr_simK3}).

@@

\newpage
\section*{Appendix}
\pagenumbering{gobble}
\renewcommand{\theequation}{A\arabic{equation}}
\setcounter{equation}{0}
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{table}{0}
\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thesection}{A\arabic{section}}
\setcounter{section}{0}

\section{Model rationale}
\subsection{Drift, admixture, and space} \label{rationale}
Here we sketch a simple model of allele frequencies and their
covariances, to justify the form given in the main text.
\paragraph{Drift} 
We first provide a simple model of allele frequencies within a layer. 
Imagine a sample $i$ that draws all of its ancestry from layer $k$.
The allele frequency in sample $i$ at locus $\ell$, $F_{i,\ell}$ can be
written as the sum
\begin{equation}
F_{i,\ell} = \epsilon_{\ell} + \Delta^{(k)}_{\ell} +
\Delta^{(k,i)}_{\ell} + \Delta^{(i)}_{\ell}
\label{drift_terms_no_admix}
\end{equation}
The first term is the ancestral allele frequency $\epsilon_\ell$ shared by all samples; 
the second is the deviation from that ancestral frequency 
due to drift in the ancestral population of the $k$th layer,
which is shared by all samples within the layer. 
The third term is the deviation of the $i$th sample away from the $k$th layer mean 
due to the spatial process of drift and migration within the layer.
The final term is the deviation specific to the $i$th sample,
which captures drift not shared by all samples at the population level
(i.e., subpopulation-specific drift due to, e.g., inbreeding). 
We will assume that these four deviations are all uncorrelated with each other.

If we have two samples $i$ and $j$ drawn from layer $k$, 
their covariance across loci will be 
\begin{equation}
\text{Var}(\epsilon) +  \text{Var}\left( \Delta^{(k)} \right) +
\text{Cov}(\Delta^{(k,i)}_{\ell},\Delta^{(k,j)}_{\ell}) + \delta_{i=j} \text{Var}\left( \Delta^{(i)} \right)\ \\
\end{equation}
where the quantity $\delta_{i=j}$ is an indicator variable that equals 1 when $i$ is equal to $j$ and 0 otherwise, 
as in Eqn. \eqref{cross_layer_covariance}.

@@

\section{Allelic covariance}\label{allelic_cov}
To see how to arrive at our empirical variance-covariance expression, Eqn. \eqref{allelic_covariance},
pick a random locus and
let $A$ and $B$ be randomly drawn alleles at that locus from populations $i$ and $j$ respectively.
If $i=j$, then these are chosen \emph{without} replacement.
Suppose these are each coded as `0' or `1` (where `0' denotes a reference allele),
but we randomly ``flip'' this coding, so that we let $X=A$ and $Y=B$ with probability 1/2,
but otherwise we let $X=1-A$ and $Y=1-B$.
We then let $\widehat{\Omega}_{i,j} = \cov[X,Y]$ ; 
sampling without replacement makes the statistic insensitive to sample size,
and random allele flipping makes it independent of the choice of reference allele.
By conditioning on the flip,
and using the fact that $\E[X] = \E[Y] = 1/2$,
Eqn. \eqref{allelic_covariance} comes from the calculation that
\begin{align}
\cov[X,Y] &= \E[XY] - 1/4 = (1/2)\E[AB + (1-A)(1-B)] - 1/4\\\notag
&= \E[(A-1/2)(B-1/2)]
\end{align}
The factor of $\delta_{i,j}/(N_i-1)$ comes from the fact that if $i=j$,
then $\E[AB]$ is the probability that both alleles drawn without replacement are reference,
which is $f_i^2 n/(n-1)$.

@@

\section{Models, parameters, and priors}\label{model_app}
\paragraph{Spatial vs. nonspatial}
In this paper, we discuss two types of models, spatial and nonspatial, 
each of which can be implemented with different numbers of layers/clusters.
The spatial model is parameterized as in Eqn \eqref{admixed_spatial_cov},
and the nonspatial model is described in Eqn \eqref{admixed_discrete_covariance}.
The nonspatial model is a special case of the spatial model, 
in which the value of all $\alpha$ parameters is set to 0.
The nonspatial model therefore has $3K$ fewer parameters than the spatial model,
as there are three $\alpha$ parameters that describe the continuous differentiation effect of distance per layer,
and they are fixed in the nonspatial model.

@@

\newpage
\section{Cross validation}\label{Xvalidation}

@@

\newpage
\section{Calculating layer contributions}\label{layer_contribution}

@@

\clearpage
\newpage
\section*{Supplementary Materials}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{equation}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

@@

\newpage

>>>>

% Template for PLoS
% Version 3.4 January 2017
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below

@@

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Inferring continuous and discrete population genetic structure across space}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Gideon S.\  Bradburd\textsuperscript{1,a},
Graham M.\  Coop\textsuperscript{2,b,\Yinyang},
Peter L.\  Ralph\textsuperscript{3,c,\Yinyang},
\\
\bigskip
\textbf{1}
Department of Integrative Biology, 
Ecology, Evolutionary Biology, and Behavior Graduate Group,
Michigan State University, MI 48824
\\
\textbf{2}
Center for Population Biology,
Department of Evolution and Ecology, 
University of California, Davis, CA 95616
\\
\textbf{3} 
Institute of Ecology and Evolution,
Departments of Mathematics and Biology,
University of Oregon, Eugene, OR 97403
\\
\textbf{a} bradburd@msu.edu; \\
\textbf{b} gmcoop@ucdavis.edu;\\
\textbf{c} plr@uoregon.edu\\

\bigskip

%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}

A classic problem in population genetics is the characterization 
of discrete population structure in the presence of 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
the use of clustering or assignment methods may incorrectly ascribe
differentiation due to continuous processes (e.g., geographic isolation by distance)
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
% This is partly a result of the difficulty of sampling uniformly and continuously 
% across the range of a population or species, but more, it 
This reflects a shortcoming of current methods for inferring and 
visualizing population structure when applied to genetic data
deriving from geographically distributed populations.
Here, we present a novel statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of two-dimensional population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This model explicitly addresses the ``clines vs. clusters" problem in 
modeling population genetic variation.
% by jointly accommodating both continuous and discrete patterns of differentiation. 
The method produces useful descriptions of structure in genetic relatedness
in situations where separated, geographically distributed populations interact,
as after a range expansion or secondary contact.
We demonstrate the utility of this approach using simulations 
and by applying it to empirical datasets of poplars and black bears in North America.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
\plr{WRITE AUTHOR SUMMARY}

\linenumbers

@@

% \section*{Methods}
\section*{Results}

@@

This definition of covariance differs from the usual ``genetic covariance'' \citep{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eq.\ \eqref{allelic_covariance} further in \secref{allelic_cov}.
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $\pi_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 \pi_{i,j})/4$.
Therefore, 
this allelic covariance is more affected in shape by singleton sites
than the standard genetic covariance,
so it may be advisable to filter these prior to analysis
if they are likely to contain a large percentage of errors.

@@

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each layer, 
we define the covariance within layer $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_layer_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th layer.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
and the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the layer.
Our choice of a powered-exponential decay, 
as parameterized by the $\alpha$s, 
is a flexible and standard choice in spatial statistics \citep{Diggle1998}, 
and is not chosen to match a particular population genetics model. 
% (Analytical theory \citep{barton2002neutral} predicts the decay should be proportional to $\exp{-D}/\sqrt{D}$.)
The $\phi^{(k)}$ is a parameter that describes the background covariance within the layer. 
If two samples draw 100\% of their ancestry from layer $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the layer.
The ``shared drift'' parameter $\phi^{(k)}$ is analogous to 
the branch length connecting the $k$th population to the population ancestral to all modeled
layers \citep[see for example ][]{patterson_ancient_2012,peter_fstats},
although they cannot be directly compared because 
we are modeling the allelic, rather than genetic, covariance. 
In \ref{rationale} we lay out a
simple model of allele frequencies underlying this covariance model.

@@

A nice feature of our approach is that the model described in Eq.\ \eqref{cross_layer_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is very similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g.][]{ADMIXTURE}; 
the \gc{main} difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomially distributed \citep{Engelhardt2012}.  
(We make this approximation for the substantial advantages in
computational speed.)
\gc{The second small difference is that in the original STRUCTURE model,
  allele frequencies at each locus are independently drawn for each
  cluster \citep{STRUCTURE}, while in conStruct's non-spatial model,
  it is more natural to envision each layer's allele frequency
  as being drifted away from a single, global allele frequency. 
  This makes our model more
  closely related to the ``$F$-model'' prior for allele frequencies of \citet{falush2003}.
As these differences are relatively small,} we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework. 
We also validate our claim that the results of our nonspatial method 
are a close match to those of STRUCTURE-based approaches.

@@

We use cross-validation \citep[similar in spirit to][]{ADMIXTURE_xval} to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} for more on our cross-validation procedure).
In general, larger values of $K$ allows the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
since above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

@@

To calculate layer contributions, we use the following alternative description of our covariance model:
the genomes of any pair of individuals \emph{agree} with some background probability,
but this probability of agreement is increased on any segment of genome 
that both have inherited through the same layer 
(the amount it increases depends on how far apart they are.)
We use this characterization to quantify the relative contributions of each layer,
by computing the average contribution to increased probability of agreement
as described in \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

% \section*{Results}

@@

To test the method, we first generated data using the coalescent simulator \texttt{ms} \citep{Hudson2002}.
In each simulation,
we split a single ancestral population into $K$ subpopulations $\tau_{\text{s}}$ units of coalescent time in the past,
and at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonized a
separate $6 \times 6$ square lattice of demes.
Migration on each lattice was to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past, 
we collapsed those $K$ discrete layers into a single grid of demes,
choosing various amounts of admixture from these different layers
(see Fig \ref{sim_setup}). 
We collapsed the layers together using random admixture with spatially autocorrelated proportions 
(see \secref{sim_details} for details), 
so that both the frequencies in layers and the layer memberships are
spatially autocorrelated. 
We simulated datasets using $K=1$, 2, and 3 layers;
in each simulation we sampled 10,000 unlinked loci from each of 20 haploid individuals from every deme.
We then ran both spatial and nonspatial \texttt{conStruct} analyses on each simulated dataset with $K$ between 1 and 7,
and compared predictive performance of the models using cross-validation.
% To confirm that results from our nonspatial model were qualitatively similar to those of 
% other nonspatial model-based clustering algorithms, 
For comparison,
we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, using $K=2-4$.
% inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}--\ref{fastStr_simK3}).

@@

\newpage
\section*{Materials and Methods}

% \section*{Appendix}
% \pagenumbering{gobble}
% \renewcommand{\theequation}{A\arabic{equation}}
% \setcounter{equation}{0}
% \renewcommand{\thetable}{A\arabic{table}}
% \setcounter{table}{0}
% \renewcommand{\thefigure}{A\arabic{figure}}
% \setcounter{figure}{0}
% \renewcommand{\thesection}{A\arabic{section}}
% \setcounter{section}{0}

\section*{Model rationale}
\subsection*{Drift, admixture, and space} \label{rationale}
Here we sketch a simple model of allele frequencies and their
covariances, to justify the form given in the main text.
\paragraph{Drift} 
We first provide a simple model of allele frequencies within a layer. 
Imagine a sample $i$ that draws all of its ancestry from layer $k$.
The allele frequency in sample $i$ at locus $\ell$, $F_{i,\ell}$ can be
written as the sum
\begin{equation}
F_{i,\ell} = \epsilon_{\ell} + \Delta^{(k)}_{\ell} +
\Delta^{(k,i)}_{\ell} + \Delta^{(i)}_{\ell}
\label{drift_terms_no_admix}
\end{equation}
The first term is the ancestral allele frequency $\epsilon_\ell$ shared by all samples; 
the second is the deviation from that ancestral frequency 
due to drift in the ancestral population of the $k$th layer,
which is shared by all samples within the layer. 
The third term is the deviation of the $i$th sample away from the $k$th layer mean 
due to the spatial process of drift and migration within the layer.
The final term is the deviation specific to the $i$th sample,
which captures drift not shared by all samples at the population level
(i.e., subpopulation-specific drift due to, e.g., inbreeding). 
We will assume that these four deviations are all uncorrelated with each other.

If we have two samples $i$ and $j$ drawn from layer $k$, 
their covariance across loci will be 
\begin{equation}
\text{Var}(\epsilon) +  \text{Var}\left( \Delta^{(k)} \right) +
\text{Cov}(\Delta^{(k,i)}_{\ell},\Delta^{(k,j)}_{\ell}) + \delta_{i=j} \text{Var}\left( \Delta^{(i)} \right)\ \\
\end{equation}
where the quantity $\delta_{i=j}$ is an indicator variable that equals 1 when $i$ is equal to $j$ and 0 otherwise, 
as in Eq.\ \eqref{cross_layer_covariance}.

@@

\section*{Allelic covariance}\label{allelic_cov}
To see how to arrive at our empirical variance-covariance expression, Eq.\ \eqref{allelic_covariance},
pick a random locus and
let $A$ and $B$ be randomly drawn alleles at that locus from populations $i$ and $j$ respectively.
If $i=j$, then these are chosen \emph{without} replacement.
Suppose these are each coded as `0' or `1` (where `0' denotes a reference allele),
but we randomly ``flip'' this coding, so that we let $X=A$ and $Y=B$ with probability 1/2,
but otherwise we let $X=1-A$ and $Y=1-B$.
We then let $\widehat{\Omega}_{i,j} = \cov[X,Y]$ ; 
sampling without replacement makes the statistic insensitive to sample size,
and random allele flipping makes it independent of the choice of reference allele.
By conditioning on the flip,
and using the fact that $\E[X] = \E[Y] = 1/2$,
Eq.\ \eqref{allelic_covariance} comes from the calculation that
\begin{align}
\cov[X,Y] &= \E[XY] - 1/4 = (1/2)\E[AB + (1-A)(1-B)] - 1/4\\\notag
&= \E[(A-1/2)(B-1/2)]
\end{align}
The factor of $\delta_{i,j}/(N_i-1)$ comes from the fact that if $i=j$,
then $\E[AB]$ is the probability that both alleles drawn without replacement are reference,
which is $f_i^2 n/(n-1)$.

@@

\section*{Models, parameters, and priors}\label{model_app}
\paragraph{Spatial vs. nonspatial}
In this paper, we discuss two types of models, spatial and nonspatial, 
each of which can be implemented with different numbers of layers/clusters.
The spatial model is parameterized as in Eq.\ \eqref{admixed_spatial_cov},
and the nonspatial model is described in Eq.\ \eqref{admixed_discrete_covariance}.
The nonspatial model is a special case of the spatial model, 
in which the value of all $\alpha$ parameters is set to 0.
The nonspatial model therefore has $3K$ fewer parameters than the spatial model,
as there are three $\alpha$ parameters that describe the continuous differentiation effect of distance per layer,
and they are fixed in the nonspatial model.

@@

\newpage
\section*{Cross validation}\label{Xvalidation}

@@

\newpage
\section*{Calculating layer contributions}\label{layer_contribution}

@@

\clearpage
\newpage
\section*{Supplementary Information}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{equation}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

@@
